Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2021-05-13 paper-revision/policar2020-tsne-embedding.tex
Add revision files

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{todo}

@@

\newcommand{\etal}{\textit{et al.}}

@@

Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data. When jointly visualising multiple data sets, a straightforward application of these methods often fails; instead of revealing underlying classes, the resulting visualizations expose dataset-specific clusters. To circumvent these batch effects, we propose an embedding procedure that uses a t-SNE visualization constructed on a reference data set as a scaffold for embedding new data points. Each data instance in the secondary data is embedded independently and does not change the reference embedding. This prevents any interactions between instances in the secondary data and implicitly mitigates batch effects. We demonstrate the utility of this approach by analyzing six recently published single-cell gene expression data sets with up to tens of thousands of cells and thousands of genes. The batch effects in our studies are particularly strong as the data comes from different institutions using different experimental protocols. The visualizations constructed by our proposed approach are clear of batch effects, and the cells from secondary data sets correctly co-cluster with cells of the same type from the primary data. We also show the predictive power of our visual classification approach in t-SNE space matches the accuracy of machine learning techniques that consider the entire compendium of features that profile single cells.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or
uniform manifold approximation and projection~\cite{umap} have recently
complemented traditional data transformation and embedding approaches such as
principal component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data from
a single coherent source, these methods may encounter problems with multiple
data sources. Here, when performing dimensionality reduction on a merged data
set, the resulting visualizations would typically reveal source-specific
clusters instead of grouping data instances of the same class, regardless of
data sources. This source-specific confounding is often referred to as {\em
domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift}
or {\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due to high biomedical relevance and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly removed. Recently
proposed procedures for batch effect removal include canonical correlation
analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
works, batch effects are deemed removed when cells from different sources
exhibit good mixing in a t-SNE visualization.  The elimination of batch effects
may require aggressive data preprocessing which may blur the boundaries between
cell types. Another problem is also the inclusion of any new data, for which
the entire analysis pipeline must be rerun, usually resulting in a different
embedding layout and clusters that have little resemblance to original visualization and
thus require reinterpretation.

@@

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, we here provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making its application
impractical for large data sets. We adopt a recent approach
for low-rank approximation of gradients based on polynomial interpolation which
reduces its time complexity to $\mathcal{O}(N)$. This approximation
enables the visualization of massive data sets, possibly containing millions of
data points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of neighbors for which
the distances in the embedding space are preserved. Small values of perplexity
result in tightly-packed clusters of points and effectively ignore the
long-range interactions between clusters. Larger values may result in a more
globally consistent visualizations --- preserving distances on a large scale
and organizing clusters in a more meaningful way --- but can lead to merging
small clusters and thus obscuring local aspects of the
data~\cite{art_of_using_tsne}.

The trade-off between the local organization and global consistency may be
achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}.
Multi-scale kernels are defined as

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. A similar formulation of
multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
resulting embeddings are visually very similar to those obtained with the
approach described above (not shown for brevity).

When using t-SNE on larger data sets, the standard learning rate $\eta = 200$
has been shown to lead to slower convergence and requires more iterations to
achieve consistent embeddings~\cite{belkina2019automated}. We follow the
recommendation of Belkina \etal ~and use a higher learning rate $\eta = N / 12$
when visualizing larger data sets.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum to speed up the
convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient
descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity needed to
evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
however, by adapting the same polynomial interpolation based approximation,
this is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding remains fixed.

@@

We apply the proposed approach to t-SNE visualizations of single-cell data.
Data in this realm include a variety of cells from specific tissues and are
characterized through gene expression.  In our experiments, we considered
several recently published data sets where cells were annotated with the cell
type. Our aim was to construct t-SNE visualizations where similarly-typed cells
would cluster together, despite systematic differences between data sources. To
that end, we focus on comparing different ways of using t-SNE rather than
differences to embeddings like PCA or MDS, which have been substantially
covered before~\cite{tsne,umap_single_cell}.  Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference: tanycytes, ependymal,
epithelial, and unlabelled cells.

\item[Human pancreas.] Baron \etal~\cite{baron2016} created
an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

\subsection{Results and Discussion}
\label{sec:results}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
  cells from the same tissue from Xin \etal~\cite{xin2016}. Just like in
  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
  secondary data set were correctly mapped to the same-typed cluster of
  reference cells.}
  \label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_retina.pdf}
  \caption{An embedding of a large reference of retinal cells from Macosco
  \etal~\cite{macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} \textbf{(b)}. We use
  colors consistent with the study by Macosko \etal}
  \label{fig:transform_retina}
\end{figure}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
9min NN-search, 13sec optimization).

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
positioning alone is insufficient. We construct a reference embedding using only
neurons from Hrvatin \etal~\cite{hrvatin2018} (Fig.~\ref{fig:optimization}.a)
and use that to position neuronal
cells from the data set from Campbell \etal~\cite{campbell2017}. We utilize the
weighted mean and median positions to initialize point positions from the secondary
data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
After initialization, we optimize point positions using the procedure described
above for 500 iterations. The resulting visualizations from both initializations
are visually very similar, indicating stable convergence. We show one of the resulting
visualizations in Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new data
  points onto reference embeddings. {\bf (a)} We construct a reference embedding
  using only neuronal subtypes from Hrvatin \etal~\cite{hrvatin2018}. {\bf (b)}
  We position neuronal cells from Campbell \etal~\cite{campbell2017} using the
  median initizaliation scheme from Kobak and Berens~\cite{art_of_using_tsne} and
  run optimization for 500 iterations. Compare the optimized embedding with the 
  initial median initialization {\bf (c)} or by using a simple weighted mean
  initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

The approach assumes that all cell types from the secondary data set are
present in the reference. The method may fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently proposed to cope
with such cases and identify the cells whose type is new and not included in
the reference~\cite{scmap}.

Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{hyperref}

@@

\newcommand{\etal}{\textit{et al.}}
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}

@@

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When jointly visualising multiple data
sets, a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
dataset-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that uses a t-SNE visualization constructed on a reference
data set as a scaffold for embedding new data points. Each data instance from a
new, unseen, secondary data is embedded independently and does not change the
reference embedding. This prevents any interactions between instances in the
secondary data and implicitly mitigates batch effects. We demonstrate the
utility of this approach by analyzing six recently published single-cell gene
expression data sets with up to tens of thousands of cells and thousands of
genes. The batch effects in our studies are particularly strong as the data
comes from different institutions using different experimental protocols. The
visualizations constructed by our proposed approach are clear of batch effects,
and the cells from secondary data sets correctly co-cluster with cells of the
same type from the primary data. We also show the predictive power of our
simple, visual classification approach in t-SNE space matches the accuracy of
specialized machine learning techniques that consider the entire compendium of
features that profile single cells.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{Maaten2008} or
uniform manifold approximation and projection~\cite{McInnes2018} have recently
complemented traditional data transformation and embedding approaches such as
principal component analysis (PCA)~\cite{wold1987principal} and multi-dimensional
scaling~\cite{cox2008multidimensional}. While useful for visualizing data from
a single coherent source, these methods may encounter problems with multiple
data sources. Here, when performing dimensionality reduction on a merged data
set, the resulting visualizations would typically reveal source-specific
clusters instead of grouping data instances of the same class, regardless of
data sources. This source-specific confounding is often referred to as {\em
domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift}
or {\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{Butler2018,Haghverdi2018,Stuart2019}.

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due to high biomedical relevance and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{Wolf2018}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{Hrvatin2018} and the
hypothalamus~\cite{Chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{Baron2016} and similarly-typed cells from
diabetic patients~\cite{Xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly removed. Recently
proposed procedures for batch effect removal include canonical correlation
analysis~\cite{Butler2018} and mutual
nearest-neighbors~\cite{Haghverdi2018,Stuart2019}.  In these works, batch
effects are deemed removed when cells from different sources exhibit good mixing
in a t-SNE visualization. The elimination of batch effects may require
aggressive data preprocessing which may blur the boundaries between cell types.
Another problem is also the inclusion of any new data, for which the entire
analysis pipeline must be rerun, usually resulting in a different embedding
layout and clusters that have little resemblance to original visualization and
thus require reinterpretation.

@@

The proposed solution implements a mapping of new data into an existing t-SNE
visualization. While the utility of such an algorithm was already hinted at in
recent publication~\cite{Kobak2019}, we here provide its practical and
theoretically-grounded implementation. Considering the abundance of recent
publications on batch effect removal, we present surprising evidence that a
computationally more direct and principled embedding procedure solves the batch
effects problem when constructing interpretable visualizations from different
data sources.

\section{Related Work}

Batch effects are systematic biases between biological data sets caused by
technical factors in the data collection and preparation process. It has been
well documented that even small differences in the experimental setup of
cell-dissociation, handling protocols, library-preparation technologies, or
sequencing platforms can significantly affect the resulting gene-expression
measurements~\cite{Tung2017,Hicks2018}. When performing downstream comparative
analyses, batch effects may confound real biological variability and may
introduce spurious correlations, leading to misleading conclusions.

Due to their severity, numerous computational approaches have been proposed to
directlty remove batch effects when performing joint analysis on two or more data sets.
Batch effect removal is typically performed as a preprocessing step. Existing
approaches involve either modifying the original data matrix or finding a joint
lower-dimensional space, where batch-effects are removed. Current methods
broadly fall into two categories: 1) mutual nearest neighbor based approaches
aim to identify matching populations of cells across the data sets, using them
to either find and correct the data sets~\cite{Haghverdi2018} or directly
construct a batch-corrected k-nearest neighbor graph used in downstream
analyses~\cite{Park2018}. 2) embedding multiple data sets into a joint
lower-dimensional space, where batch effects are removed. Some of these
approaches opt for linear dimensionality-reduction methods such as
PCA~\cite{Korsunsky2019} or MultiCCA~\cite{Butler2018}, while others employ
non-linear techniques from deep learning~\cite{Li2020,Lopez2018} Still, other
approaches use a combination of the two~\cite{Stuart2019,Hie2019}.

Orthogonal approaches aimed at cell-type attempt to circumvent batch effects
altogether by identifying a subset of representative genes which act as markers
for a specific cell-type. Instead of considering the entire gene-expression
profile, which may be noisy and affected by batch effects, the idea is instead
to inspect the cells for the presence of a handful of genes, which could
adequately be used to determine the cell type. One such procedures is
scMap-Cluster, a consensus-based KNN method, tailored explicitly to scRNA-seq
gene-expression data~\cite{Kiselev2018}. scMap-Cluster uses three
correlation-based distance measures and uses a voting scheme to perform
classification. To identify novel cell types, scMap-Cluster heuristically
determines a distance threshold.

Broadly speaking, our approach falls into the second category, alongside
scMap-Cluster. The approach we describe uses standard single-cell pipelines
based on identifying representative genes. However, we emphasize that the
primary purpose of our approach is not classification, but the visualization of
the various cell-types. We can apply a KNN classifier to the resulting
visualizations to obtain accuracy estimates and compare our approach to other
classification methods. However, the classification aspect of our approach is
secondary: the primary purpose of t-SNE is to aid in scientists in exploratory
data analysis and help them better understand the underlying data landscape.

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making its application
impractical for large data sets. We adopt a recent approach for low-rank
approximation of gradients based on polynomial interpolation which reduces its
time complexity to $\mathcal{O}(N)$. This approximation enables the
visualization of massive data sets, possibly containing millions of data
points~\cite{Linderman2019}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of neighbors for which
the distances in the embedding space are preserved. Small values of perplexity
result in tightly-packed clusters of points and effectively ignore the
long-range interactions between clusters. Larger values may result in a more
globally consistent visualizations --- preserving distances on a large scale and
organizing clusters in a more meaningful way --- but can lead to merging small
clusters and thus obscuring local aspects of the data~\cite{Kobak2019}.

The trade-off between the local organization and global consistency may be
achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
a mixture of Gaussians of varying bandwidths~\cite{Lee2015}. Multi-scale kernels
are defined as

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. A similar formulation of multi-scale kernels
was proposed in~\cite{Kobak2019}, and we found the resulting embeddings are
visually very similar to those obtained with the approach described above (not
shown for brevity).

When using t-SNE on larger data sets, the standard learning rate $\eta = 200$
has been shown to lead to slower convergence and requires more iterations to
achieve consistent embeddings~\cite{Belkina2019}. We follow the recommendation
of Belkina \etal ~and use a higher learning rate $\eta = N / 12$ when
visualizing larger data sets.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum to speed up the
convergence, as proposed by Jacobs~\cite{Jacobs1988,Maaten2014}. We run gradient
descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity needed to
evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
however, by adapting the same polynomial interpolation based approximation, this
is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding
remains fixed.

@@

We apply the proposed approach to t-SNE visualizations of single-cell data. Data
in this realm include a variety of cells from specific tissues and are
characterized through gene expression.  In our experiments, we considered
several recently published data sets where cells were annotated with the cell
type. Our aim was to construct t-SNE visualizations where similarly-typed cells
would cluster together, despite systematic differences between data sources. To
that end, we focus on comparing different ways of using t-SNE rather than
differences to embeddings like PCA or MDS, which have been substantially covered
before~\cite{Maaten2008,Becht2019}. Below, we list the data sets used in our
experiments, and display the resulting data visualizations. Due to the unique
nature of single-cell data, we apply a specialized single-cell pipeline for all
our experiments, as described in Appendix~\ref{sec:sc-pipeline}. Finally, we
discuss the success of the proposed approach in alleviating the batch effects.

@@

\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{Hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{Chen2017}, containing cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference: tanycytes, ependymal,
epithelial, and unlabelled cells.

\item[Human pancreas.] Baron \etal~\cite{Baron2016} created
an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{Xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] Macosko \etal~\cite{Macosko2015} created
an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{Shekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

\subsection{t-SNE transform successfully alleviates batch effects}
\label{sec:results}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
  \caption{Embedding of pancreatic cells from Baron \etal~\cite{Baron2016} and
  cells from the same tissue from Xin \etal~\cite{Xin2016}. Just like in
  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
  secondary data set were correctly mapped to the same-typed cluster of
  reference cells.}
  \label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_retina.pdf}
  \caption{An embedding of a large reference of retinal cells from Macosco
  \etal~\cite{Macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
  focuses on bipolar cells from Shekhar \etal~\cite{Shekhar2016} \textbf{(b)}. We use
  colors consistent with the study by Macosko \etal}
  \label{fig:transform_retina}
\end{figure}

\subsection{Constructing a reference embedding}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

@@

\subsection{Optimization is crucial to producing meaningful point embeddings}

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{Kobak2019}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
9min NN-search, 13sec optimization).

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
positioning alone is insufficient. We construct a reference embedding using only
neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a)
and use that to position neuronal
cells from the data set from Campbell \etal~\cite{Campbell2017}. We utilize the
weighted mean and median positions to initialize point positions from the secondary
data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
After initialization, we optimize point positions using the procedure described
above for 500 iterations. The resulting visualizations from both initializations
are visually very similar, indicating stable convergence. We show one of the resulting
visualizations in Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new data
  points onto reference embeddings. {\bf (a)} We construct a reference embedding
  using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)}
  We position neuronal cells from Campbell \etal~\cite{Campbell2017} using the
  median initizaliation scheme from Kobak and Berens~\cite{Kobak2019} and
  run optimization for 500 iterations. Compare the optimized embedding with the 
  initial median initialization {\bf (c)} or by using a simple weighted mean
  initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

\subsection{Our approach requires a complete reference set}

Our approach assumes that all cell types from the secondary data set are present
in the reference. Intuitively, using t-SNE in such a way is conceptually similar
to classification via $k$-nearest neighbor classifiers and is similarily
limited. The method may fail to reveal unseen cell types in the secondary data
set, likely positioning them arbitrarily close to unrelated clusters. In some
instances, unknown cell types may be sufficiently different from the reference
data that t-SNE will repel them from existing clusters. However, we caution that
this approach is unreliable and depends heavily on the chosen preprocessing
pipeline.

We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
create a reference embedding containing only neuronal cells from Hrvatin
\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
\etal~\cite{Campbell2017} and add them to the reference embedding in
Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal
are scattered somewhat arbitrarily around several clusters in the reference
embedding. Interestingly, the secondary data points form a ``ring'' around one
of the clusters, indicating that these data points are very different from the
cells in this cluster. Notice also that the points from the secondary data set
exhibit little to no clustering and the different cell types seem to be mixed
amongst each other. We hypothesize that this effect is due primarily to the
single-cell preprocessing pipeline and not the limitations of our proceedure
itself, as the informative genes selected to create the reference neuronal
embedding likely do not differentiate supportive glial cells from the secondary
data set. This effect is similar to procedures such as scMap-Cluster, a consensus KNN
method, which heuristically determines a distance threshold to identify unknown
cell types~\cite{Kiselev2018}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
  \caption{A reference embedding must contain all the cell-types in the
  secondary embedding to produce reliable results. {\bf (a)} We construct a
  reference embedding containing only neuronal cells from Hrvatin
  \etal~\cite{Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from
  Campbell \etal~\cite{Campbell2017} so that no overlap exists between the
  cell-types between the data sets. Conceptually, t-SNE behaves similarily to a
  KNN classifier, and places the non-neurnal cells to their the most similar
  points in the reference. In some instances, the non-neuronal cells are
  sufficiently different from the neuronal cells so that they are repelled from
  the reference clusters. Such behaviour results in the ``ring'' seen on the
  right hand side of the embedding.} \label{fig:incomplete_reference}
\end{figure}

Our procedure is, therefore, asymmetrical in the choice of reference and
secondary data set. In practice, however, newly produced {\em secondary} data
would be embedded into previously-prepared reference landscapes. Large
collections of data {\em e.g.} the Human Cell Atlas
initiative~\cite{Rozenblatt-Rosen2017} make it possible to scale up our approach
to wider sets of cell types. Identifying potential failure cases where rare
cell-types may still be missing from constructed reference embeddings is a
problem that plagues the bioinformatics community and is an active area of
research.

\subsection{Comparison to other similar batch-effect methods}
